Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-10-31 paper/paper.tex
thompson wip

\paragraph{Regex-NFA/Thompson}

\steven{TODO put regex-NFA iso here. Call it Thompson's construction?. Mention
  its mechanized}

@@

\subsection{The Dyck Grammar}
%% 3. Dyck language
The Dyck grammar $D$ for balanced parens (summarize what it is + citation)

\steven{somewhere in here, sneak in the benchmark where the Dyck parser could
  handle 10's of thousands of characters in about 30s}

\steven{this is a VERY loose sketch of the things I want to say about $D$}
\Cref{fig:DyckAutomaton} presents an automaton $M$ that recognizes $D$. We show this
precisely by constructing an isomorphism between $D$ and $\Trace_{M}~0~\true$.
Unlike the finite automata seen thus far, $M$ has infinitely many states --- one
for each $n : \mathbb{N}$ with an additional fail state. This diverges a bit
from classical language theory, where the recognizer for $D$ would be encoded as a
pushdown automaton with finitely many
states and stack of memory. Such PDAs are encodable in our theory, although they
are a little cumbersome to use. Moreover, they
seem overfit to classical reasoning as they don't preserve the grammar up to
strong equivalence, only weak equivalence in general. \steven{cite Lambek
  automata and state that is is a secondary goal to preserve as much structure as possible}
By moving the stack of memory (the nats) into the
state, we obtain infinitely many states, which is fine bc the states need only
be a type for us to interface with it.

\steven{TODO sketch proof of equivalence. Mention that it is mechanized}

\begin{figure}
  \begin{tikzpicture}[node distance = 17mm ]
    \node[state, initial, accepting] (0) {$0$};
    \node[state, right of=0] (1) {$1$};
    \node[state, right of=1] (2) {$2$};
    \node[right of=2] (3) {$\dots$};
    \node[state, below of=0] (fail) {$\mathsf{fail}$};

    \path[->] (0) edge[above, bend left] node{\stringquote{(}} (1)
              (1) edge[below, bend left] node{\stringquote{)}} (0)
              (1) edge[above, bend left] node{\stringquote{(}} (2)
              (2) edge[below, bend left] node{\stringquote{)}} (1)
              (2) edge[above, bend left] node{\stringquote{(}} (3)
              (3) edge[below, bend left] node{\stringquote{)}} (2)
              (fail) edge[loop right] node{\stringquote{(}, \stringquote{)}} (fail)
              (0) edge[left] node{\stringquote{)}} (fail);
  \end{tikzpicture}
  \caption{Automaton $M$ for the Dyck grammar}
  \label{fig:DyckAutomaton}
\end{figure}

>>>>

@@

\begin{theorem}[Thompson's Construction, \Agda]
  Given a regular expression $R$, there exists an NFA $N$ such that $R$ is
  strongly equivalent to $\Trace_{N}(N.\init)$.
\end{theorem}

\begin{proof}
In this proof, we present a variant of Thompson's construction
\cite{thompsonProgrammingTechniquesRegular1968} in which we induct on the
structure of $R$ to construct strongly equivalent NFAs. We will sketch this
proof by walking through some of the inductive cases, the remaining cases follow
similar reasoning and the details may be found in the formalized artifact.

\begin{figure}
  \begin{tikzpicture}[node distance = 17mm ]
    \node[state, initial] (0) {$0$};
    \node[state, right of=0, accepting] (1) {$1$};

    \path[->] (0) edge[above] node{$c$} (1);
  \end{tikzpicture}
  \caption{NFA $N_{c}$ for a single character $c$}
  \label{fig:litnfa}
\end{figure}

First handle the case where $R$ is a literal character $c$. The NFA $N_{c}$ pictured in
\cref{fig:litnfa} is strongly equivalent to the grammar $c$. To demonstrate this
strong equivalence, we build terms between $c$ and $\Trace_{N_{c}}(N_{c}.\init)$
and show they are mutually inverse.

Define $f : \ltonl {(c \lto \Trace_{N_{c}}(N_{c}.\init))}$ as follows,

\[
  a : c \vdash f := \cons(a, \nil) : \Trace_{N_{c}}(N_{c}.\init)
\]

Define $g : \ltonl {(\Trace_{N_{c}}(N_{c}.\init) \lto c)}$ via induction on the
trace. Map traces beginning in state 1, which are necessarily built via $\nil$,
to the linear unit $I$. Then we must handle traces beginning in state 0, which
are necessarily built via $\cons$. We decompose such a trace into a parse of the
character $c$ and a trace beginning in state 1. We inductively transform the
trace beginning in state 1 into a term of type $I$. Finally, we use the right
unitor to turn a term of type $c \otimes I$ into one of type $c$.

\[\begin{tikzcd}
    {c \otimes \mathsf{Trace}_{N_{c}}(1)} &
    {c \otimes I} &
    c
	\arrow["{id\otimes g}", from=1-1, to=1-2]
	\arrow["{}", from=1-2, to=1-3]
\end{tikzcd}\]

Note that the grammar $c$ is unambiguous, so $g \circ f$ is equal to the
identity on $c$. Additionally, we can show that $f$ carries the structure of an
indexed algebra homomorphism for the family of functors defining the trace type
$\Trace_{N_{c}}$. Because $g$ is defined via induction, it carries the structure
of an algebra homomorphism as well. Thus, $f \circ g$ is also an algebra
homomorphism. $f \circ g$ and the identity map are both then algebra
homomorphisms from $\Trace_{N_{c}}$ to itself. Because $\Trace_{N_{c}}$ is the
initial algebra for the relevant family of functors, there is only one
homomorphism from it to itself and thus $f \circ g = \mathsf{id}$. Therefore,
$f$ and $g$ are mutually inverse and the regular expression is strongly
equivalent to the traces through the NFA in this case.

Now consider the case where $R$ is the disjunction of two regular expressions
$R_{1}$ and $R_{2}$. Inductively, we may build NFAs $N_{1}$ and $N_{2}$ that are
strongly equivalent to $R_{1}$ and $R_{2}$, respectively. We define $N_{\oplus}$
as a new NFA built as the disjunction of $N_{1}$ and $N_{2}$.
$N_{\oplus}.\states$ comprises the states from $N_{1}$, the states from
$N_{2}$, and a new marked initial state. The accepting states are precisely
those that were accepting in the subautomata. The initial state has an
$\epsilon$-transition to each of the initial states of the subautomata. Once in
$N_{1}.\init$, the downstream transitions are a copy of those in $N_{1}$,
likewise for $N_{2}.\init$.  That is, $N_{\oplus}$ contains a copy of each
$N_{1}$ and $N_{2}$, and we choose which copy to inhabit by deciding which
$\epsilon$-transition out of the initial state to take. A schematic
representation of this automaton is given in \cref{fig:disjunctionnfa}.

\begin{figure}
  \begin{tikzpicture}[node distance = 17mm ]
    \node[state, initial] (init) {$\init$};
    \node[state, above right of=init] (init1) {$N_{1}.\init$};
    \node[state, below right of=init] (init2) {$N_{2}.\init$};
    \node[right of=init1] (dots1) {$\dots$};
    \node[right of=init2] (dots2) {$\dots$};
    \node[state, accepting, right of=dots1] (acc1) {$s_{1}$};
    \node[state, accepting, right of=dots2] (acc2) {$s_{2}$};

    \path[->] (init) edge[below] node{$\epsilon$} (init1)
              (init) edge[above] node{$\epsilon$} (init2)
              (init1) edge[above] node{} (dots1)
              (init2) edge[above] node{} (dots2)
              (dots1) edge[above] node{} (acc1)
              (dots2) edge[above] node{} (acc2);
  \end{tikzpicture}
  \caption{NFA $N_{\oplus}$ as a disjunction of $N_{1}$ and $N_{2}$}
  \label{fig:disjunctionnfa}
\end{figure}

Because $N_{1}$ and $N_{2}$ are each strongly equivalent to $R_{1}$ and $R_{2}$,
it suffices to show that
$\Trace_{N_{1}}(N_{1}.\init) \oplus \Trace_{N_{2}}(N_{2}.\init)$ is strongly
equivalent to $\Trace_{N_{\oplus}}(N_{\oplus}.\init)$.

Just like the case of a single character, we induct on traces when defining this
strong equivalence. First we map out of $\Trace_{N_{\oplus}}(s)$. If $s$ is a
state from $N_{1}$, then we send the trace from $N_{\oplus}$

\steven{TODO finish this, my brain isn't working rn}

\end{proof}
